Supermartingales and Termination Analysis of Statistical PCF

### 2020-08-26 probabilisticProgrammingMartingales.tex
minor

A position \akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
\end{align*}
The \changed[lo]{\emph{subterm of $M$ at a position $P$}, denoted $M \mid P$,} is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; P & = M \mid P \\
M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
\tY M \mid \tY ; P & = M \mid P \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; P & = M \mid P
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
A position such that $M\mid P$ does exist is said to occur in $M$. 
\changed[lo]{Substitution of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
For example 
\begin{align*}
M &= \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})\\
P &=\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\end{align*} 
then 
\(
M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)}

@@

\paragraph{}
For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
\begin{itemize}
    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ \lo{Define $\sim_N^*$.}
    \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
    \item If $M \to N$ at $P$, no position in any term in $Rch(N)$ is related to $(M,P)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
    \item The subprobability distribution of values obtained by reducing a term in the usual call-by-value semantics is equal to that obtained by any reduction strategy on $\Rightarrow$ (i.e.~a sub-relation of $\Rightarrow$ that is a function) (with some allowance made for the fact that some values are not normal forms of $\Rightarrow$, so the reduction strategy might overshoot the value that $\red^\infty$ reaches. I'm not quite sure what form this will take.)
    \item Similarly, the probability of termination is equal in $\red^\infty$ and any reduction strategy on $\Rightarrow$, so almost sure termination can be equivalently defined based on $\Rightarrow$.
    \item For any closed term $N$, if there is any reduction strategy for $N$ and ranking function corresponding to that reduction strategy, $N$ is AST.
    \item Corresponding sparse and antitone-strict ranking function theorems.
\end{itemize}

@@

\begin{proof}
Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur. It is only required that the reductions that do occur occur in the correct order.
\lo{This paragraph (especially the last sentence) is ambiguous.
What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
\lo{So this is permitted?
\[
(II)(I (I \bar I)) \to
(II)(I\bar I) \to
(II)\bar I 
\]
but not this: 
\(
(II)(I (I \bar I)) \to
(II)(I \bar I) \to
(II)\bar I 
\)?}

@@

\akr{For the standard CBV semantics, this example is indeed not a problem. 
In the alternative semantics, defining 
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
The set of samples where it terminates is 
\[\begin{array}{l}
\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
}_1;@_1;\mathsf{if}_1) = r, \\
\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
\end{array}\] 
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
It is this set that's unmeasurable.}

>>>>

A \emph{position} 
\akr{I'm not really sure how to phrase the citation, but this is adapted from \cite{Kennaway96infinitarylambda}.} 
\lo{The idea is common enough. I think no citation necessary.}
is a finite sequence of steps into a term, defined inductively as
\begin{align*}
P ::= \cdot \mid \lambda ; P \mid @_1 ; P \mid @_2 ; P \mid \underline f_i ; P \mid \tY ; P \mid \textsf{if}_1 ; P \mid \textsf{if}_2 ; P \mid \textsf{if}_3 ; P \mid \tscore ; P.
\end{align*}
The \emph{subterm of $M$ at a position $P$}, denoted $M \mid P$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; P & = M \mid P \\
M_1 M_2 \mid @_i ; P & = M_i \mid P \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; P & = M_i \mid P \quad \text{for }i \leq n \\
\tY M \mid \tY ; P & = M \mid P \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; P & = M_i \mid P \quad \text{for } i = 1,2,3 \\
\tscore(M) \mid \tscore ; P & = M \mid P
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda ; \cdot$ is undefined). 
A position such that $M\mid P$ does exist is said to occur in $M$. 
\changed[lo]{\emph{Substitution} of $N$ at position $P$ in $M$, written $M[N/P]$, is defined similarly.
For example 
\begin{align*}
M &= \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})\\
P &=\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\end{align*} 
then 
\(
M[\tsample / P] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)}

@@

\paragraph{}
For now, here's a summary of the lemmas and theorems I intend to prove about this alternative semantics:
\begin{itemize}
    \item If $M \to N$, the relation $\sim_N^*$ defined on $Rch(N)$ is equal to the restriction of $\sim_M^*$ to $Rch(N)$ %\lo{Define $\sim_N^*$.}
    \item Given the definition below of $\Rightarrow$, which is still applicable to this version of $\sim$, and which is the equivalent of $\red$ in this version of the semantics, $\Rightarrow$ is Church-Rosser.
    \item If $M \to N$ at $P$, no position in any term in $Rch(N)$ is related to $(M,P)$ by $\sim_M^*$, and therefore each sample is used at most once in any reduction sequence.
    \item The subprobability distribution of values obtained by reducing a term in the usual call-by-value semantics is equal to that obtained by any reduction strategy on $\Rightarrow$ (i.e.~a sub-relation of $\Rightarrow$ that is a function) (with some allowance made for the fact that some values are not normal forms of $\Rightarrow$, so the reduction strategy might overshoot the value that $\red^\infty$ reaches. I'm not quite sure what form this will take.)
    \item Similarly, the probability of termination is equal in $\red^\infty$ and any reduction strategy on $\Rightarrow$, so almost sure termination can be equivalently defined based on $\Rightarrow$.
    \item For any closed term $N$, if there is any reduction strategy for $N$ and ranking function corresponding to that reduction strategy, $N$ is AST.
    \item Corresponding sparse and antitone-strict ranking function theorems.
\end{itemize}

@@

\begin{proof}
Note that although the reductions that occur in $X_{cbv}$ are in call-by-value order, it may not be the actual call-by-value reduction sequence because some redexes may remain un-reduced even when other reductions that should happen afterwards occur. It is only required that the reductions that do occur occur in the correct order.
\lo{This paragraph (especially the last sentence) is ambiguous.
What does it mean to say that that a reductions that occur in a reduction sequences are \emph{in CBV order}?}
\iffalse
\lo{So this is permitted?
\[
(II)(I (I \bar I)) \to
(II)(I\bar I) \to
(II)\bar I 
\]
but not this: 
\(
(II)(I (I \bar I)) \to
(II)(I \bar I) \to
(II)\bar I 
\)?}
\fi

@@

\lo{Give an example or two of a reduction strategy (for $\Rightarrow$) that is not $\cbv$.}

\akr{NOTE. For the standard CBV semantics, this example is indeed not a problem. 
In the alternative semantics, defining 
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
The set of samples where it terminates is 
\[\begin{array}{l}
\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid s([A[\tsample]], \mathsf{if}_1;\underline{ -
}_1;@_1;\mathsf{if}_1) = r, \\
\qquad \qquad \hfill \qquad \qquad s([A[\tsample], A[r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
\end{array}\] 
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
It is this set that's unmeasurable.}
